---
title: "Longitudinal exploration of cars"
output: html_notebook
---

```{r}
library(haven)
library(tidyverse)
library(skimr)
library(corrplot)
library(scales)
library(ggthemes)
library(kableExtra)

current_dir <- getwd()
source(file.path(current_dir,"src","R","utils.R"))

years <- c(2016, 2017, 2018, 2019, 2020, 2021, 2023, 2025)

waves <- list()

data_path <- "/home/amy/tmp/sheds_data/files/SHEDS"

waves <- list()
for (year in years) {
  file_path <- paste0("/home/amy/tmp/sheds_data/files/SHEDS", year, ".sav")
  wave_name <- as.character(year)
  waves[[wave_name]] <- read_clean_sheds(file_path)
}

all_ids <- waves %>%
  map_df(~ select(.x, id)) %>%
  pull(id) %>%
  unique()

cat("\nTotal unique IDs across all waves:", length(all_ids), "\n")

cat("\n=== Summary ===\n")
cat("Total respondents (all waves):", sum(map_int(waves, nrow)), "\n")
cat("Total unique IDs:", length(all_ids), "\n")
cat("Duplicate appearances:", sum(map_int(waves, nrow)) - length(all_ids), "\n")

id_counts <- waves %>%
  map_df(~ select(.x, id), .id = "wave") %>%
  count(id) %>%
  arrange(desc(n))

cat("\nDistribution of appearances:\n")
print(table(id_counts$n))
```

EV ownership over time using md_708 and history building:
```{r}
library(tidyverse)
library(scales)
library(knitr)

md_708_summary <- bind_rows(
  lapply(names(waves), function(wave_name) {
    df <- waves[[wave_name]]
    tibble(
      year           = as.integer(wave_name),
      n_total        = nrow(df),
      n_population   = if ("md_hhgr" %in% names(df)) sum(df$md_hhgr[df$mob2_1 >= 1 & df$mob2_1 < 90], na.rm = TRUE) else NA_integer_,
      n_ev_md708     = if ("md_708" %in% names(df)) sum(df$md_708 == 1, na.rm = TRUE) else NA_integer_,
      n_ev_missing   = if ("md_708" %in% names(df)) sum(is.na(df$md_708)) else NA_integer_,
      ev_rate_md708  = n_ev_md708 / n_total
    )
  })
) %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025))

md_708_summary %>%
  mutate(ev_rate_md708 = percent(ev_rate_md708, accuracy = 0.01)) %>%
  kable(
    col.names  = c("Year", "N Total", "N Pop.", "N EV (md_708)", "N Missing", "EV Rate"),
    align       = "r",
    format.args = list(big.mark = ",")
  ) %>%
  print()

stata_car_history <- build_car_history(waves)


results_stata <- bind_rows(
  analyze_ev_ownership_data(stata_car_history, 2019, waves),
  analyze_ev_ownership_data(stata_car_history, 2020, waves ),
  analyze_ev_ownership_data(stata_car_history, 2021, waves),
  analyze_ev_ownership_data(stata_car_history, 2023, waves),
  analyze_ev_ownership_data(stata_car_history, 2025, waves)
)

results_stata %>%
  select(year, n_total, n_car_owner_population, n_car_owners, n_ev_main, n_ev_secondary,
         n_ev_total, n_hybrid_gas, n_hybrid_diesel, ev_rate_car_owners, n_car_owners_inter) %>%
  mutate(ev_rate_car_owners = percent(ev_rate_car_owners, accuracy = 0.01)) %>%
  kable(
    col.names = c("Year", "N", "Pop.", "Car Own.", "EV Main", "EV 2nd",
                  "EV Total", "Hybrid Gas", "Hybrid Dies.", "EV Rate", "N car owners inter"),
    align = "r",
    format.args = list(big.mark = ",")
  ) %>%
  print()
```


Here we want to see who are the people which have md_708 and who have ev in the car history based on the ids what are the differences.
Did people responded differently to each questions? or maybe they didnt say the changed vehicles can we track ids from both dataframes and check
why it differs?

Diagnostic: Who is in md_708 vs car history - and why do they differ?
```{r}
# For each year, build a per-respondent comparison between the two EV sources
diagnose_ev_discrepancy <- function(yr) {
  raw <- waves[[as.character(yr)]] %>%
    mutate(across(everything(), as.vector)) %>%
    select(id,
           any_of("md_708"),
           any_of("mob2_1"),
           any_of("mob3_3"),
           any_of("mob3_change"),
           any_of("mob2_e")) %>%
    mutate(
      md_708      = ifelse("md_708" %in% names(.) & !is.na(md_708) & md_708 %in% c(-1,-2), NA_real_, as.numeric(md_708)),
      mob3_3      = ifelse(mob3_3 %in% c(-2,-1), NA_real_, as.numeric(mob3_3)),
      mob3_change = ifelse("mob3_change" %in% names(.) & !is.na(mob3_change) & mob3_change %in% c(-1,-2), NA_real_, as.numeric(mob3_change)),
      mob2_e      = ifelse("mob2_e" %in% names(.) & !is.na(mob2_e) & mob2_e %in% c(-1,-2), NA_real_, as.numeric(mob2_e))
    )

hist_yr <- stata_car_history %>%
    filter(year_wave == yr) %>%
    mutate(has_ev = as.integer(mob3_3_filled == 8 | mob2_e_filled == 1)) %>%
    select(id, mob3_3_filled, mob2_e_filled, has_ev)

  joined <- raw %>%
    left_join(hist_yr, by = "id") %>%
    mutate(
      ev_md708   = as.integer(!is.na(md_708) & md_708 == 1),
      ev_history = as.integer(!is.na(has_ev) & has_ev == 1),
      group = case_when(
        ev_md708 == 1 & ev_history == 1  ~ "Both agree: EV",
        ev_md708 == 0 & ev_history == 0  ~ "Both agree: No EV",
        ev_md708 == 1 & ev_history == 0  ~ "md_708 only (history missed)",
        ev_md708 == 0 & ev_history == 1  ~ "History only (md_708 missed)",
        is.na(ev_md708) & ev_history == 1 ~ "History only (md_708 NA)",
        ev_md708 == 1 & is.na(ev_history) ~ "md_708 only (not in history)",
        TRUE                                ~ "Both NA / unclear"
      )
    )
  joined
}

diag_table <- bind_rows(
  lapply(c(2019, 2020, 2021, 2023, 2025), function(yr) {
    d <- diagnose_ev_discrepancy(yr)
    n_total <- nrow(waves[[as.character(yr)]])
    bind_rows(
      d %>% count(group) %>% mutate(year = yr, n_total = n_total),
      tibble(
        group = c(
          "TOTAL: EV in md_708 (all IDs)",
          "TOTAL: EV in history (all IDs)",
          "TOTAL: EV in either source (union)",
          "TOTAL: EV in both sources (intersection)",
          "TOTAL: md_708 missing & EV via secondary only",
          "TOTAL: md_708 missing & EV via main only"
        ),
        n = c(
          sum(d$ev_md708   == 1, na.rm = TRUE),
          sum(d$ev_history == 1, na.rm = TRUE),
          sum(d$ev_md708 == 1 | d$ev_history == 1, na.rm = TRUE),
          sum(d$ev_md708 == 1 & d$ev_history == 1, na.rm = TRUE),
          sum(is.na(d$md_708) & d$mob2_e_filled == 1 & d$mob3_3_filled != 8, na.rm = TRUE),
          sum(is.na(d$md_708) & d$mob3_3_filled == 8, na.rm = TRUE)
        ),
        year    = yr,
        n_total = n_total
      )
    )
  })
) %>%
  mutate(label = sprintf("%d (%.1f%%)", n, n / n_total * 100)) %>%
  select(group, year, label) %>%
  pivot_wider(names_from = year, values_from = label, values_fill = "0 (0.0%)") %>%
  arrange(group)

total_rows <- which(str_detect(diag_table$group, "TOTAL"))

diag_table %>%
  kable(caption = "Agreement between md_708 and car history by year (joined on IDs)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  row_spec(total_rows, bold = TRUE, background = "#eeeeee")
```


People history says EV but not md708:
```{r}
# People md_708 says has EV but history missed
diag_2025 <- diagnose_ev_discrepancy(2025)
md708_only <- diag_2025 %>%
  filter(group == "md_708 only (history missed)") %>%
  select(id, md_708, mob3_3, mob3_3_filled, mob3_change, mob2_e, mob2_e_filled)

cat("\n--- md_708 says EV, history missed:", nrow(md708_only), "respondents ---\n")
cat("Did they report a car change (mob3_change)?\n")
print(table(md708_only$mob3_change, useNA = "always"))
cat("\nRaw mob3_3 engine type in this wave:\n")
print(table(md708_only$mob3_3, useNA = "always"))
cat("\nFilled mob3_3 from history:\n")
print(table(md708_only$mob3_3_filled, useNA = "always"))
cat("\nNA mob3_3 (never reported engine type):", sum(is.na(md708_only$mob3_3)), "\n")
cat("Non-EV engine type in mob3_3 (possible misclassification):",
    sum(!is.na(md708_only$mob3_3) & !md708_only$mob3_3 %in% 8, na.rm = TRUE), "\n")
```


Prepare data to be compared with FSO:
```{r}
# SHEDS md_708: divide by n_car_owners
sheds_md708_pop <- md_708_summary %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  left_join(results_stata %>% select(year, n_car_owners), by = "year") %>%
  mutate(
    percentage = (n_ev_md708 / n_car_owners) * 100,
    type       = "Pure Electric",
    source     = "SHEDS (md_708)"
  ) %>%
  select(year, type, percentage, source)

# SHEDS history: divide by n_car_owners, include hybrids
sheds_history_pop <- results_stata %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(n_hybrid_total = n_hybrid_gas + n_hybrid_diesel) %>%
  select(year, n_ev_main, n_ev_secondary, n_hybrid_total, n_car_owners) %>%
  pivot_longer(cols = c(n_ev_main, n_ev_secondary, n_hybrid_total),
               names_to = "type", values_to = "count") %>%
  mutate(
    percentage = (count / n_car_owners) * 100,
    type       = recode(type,
                        "n_ev_main"      = "Pure Electric (Main)",
                        "n_ev_secondary" = "Pure Electric (Secondary)",
                        "n_hybrid_total" = "Hybrid"),
    source     = "SHEDS (History)"
  ) %>%
  select(year, type, percentage, source)

print(sheds_md708_pop)
print(sheds_history_pop)
```

Old BFS file:
```{r}
bfs_percentage <- read_csv("/home/amy/tmp/repos/sheds_data_analysis/gr-e-11.03.02.01-03.csv",
                           show_col_types = FALSE) %>%
  rename(year = `Unnamed: 0`) %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(
    hybrid_total   = `conventional hybrid` + `plug-in hybrid`,
    total_vehicles = petrol + diesel + `conventional hybrid` + `plug-in hybrid` + electric + other,
    ev_pct         = (electric      / total_vehicles) * 100,
    hybrid_pct     = (hybrid_total  / total_vehicles) * 100
  ) %>%
  select(year, ev_pct, hybrid_pct) %>%
  pivot_longer(-year, names_to = "type", values_to = "percentage") %>%
  mutate(
    type   = recode(type, "ev_pct" = "Pure Electric", "hybrid_pct" = "Hybrid"),
    source = "FSO (Official)"
  )

# Bind and stack it
comparison_data_population <- bind_rows(bfs_percentage, sheds_md708_pop, sheds_history_pop) %>%
  mutate(
    source = factor(source, levels = c("FSO (Official)", "SHEDS (md_708)", "SHEDS (History)")),
    type   = factor(type,   levels = c("Hybrid", "Pure Electric", "Pure Electric (Main)", "Pure Electric (Secondary)"))
  )

comparison_totals_population <- comparison_data_population %>%
  group_by(year, source) %>%
  summarise(total_pct = sum(percentage), .groups = "drop")


p_comparison_population <- comparison_data_population %>%
  left_join(comparison_totals_population, by = c("year", "source")) %>%
  ggplot(aes(x = source, y = percentage, fill = type)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
            position = position_stack(vjust = 0.5),
            size = 3.5, color = "white", fontface = "bold") +
  geom_text(aes(y = total_pct, label = sprintf("%.1f%%", total_pct)),
            vjust = -0.5, size = 3.5, fontface = "bold", color = "black") +
  scale_fill_manual(values = c(
    "Hybrid"                    = "#FF9800",
    "Pure Electric"             = "#1B5E20",
    "Pure Electric (Main)"      = "#1B5E20",
    "Pure Electric (Secondary)" = "#81C784"
  )) +
  scale_y_continuous(
    limits = c(0, max(comparison_totals_population$total_pct) * 1.15),
    breaks = seq(0, 20, 2),
    expand = c(0, 0)
  ) +
  facet_wrap(~ year, nrow = 1) +
  labs(x = NULL, y = "Percentage (%)", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title         = element_text(face = "bold", size = 16),
    axis.title.y       = element_text(size = 10, face = "bold"),
    axis.text.x        = element_text(size = 9, angle = 15, hjust = 1),
    legend.position    = "bottom",
    legend.text        = element_text(size = 16),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    strip.text         = element_text(face = "bold", size = 16)
  )

print(p_comparison_population)
save_plot(p_comparison_population, "bfs_vs_sheds_comparison_population", width = 14, height = 5)
```

New bfs file:
```{r}
fso <- read_csv("/home/amy/tmp/repos/sheds_data_scripts/src/fso_stock_vehicles.csv") %>%
  mutate(
    phev        = phev_petrol + phev_diesel,
    hev         = petrol_hev + diesel_hev,
    ev_pct      = (bev  / total) * 100,
    phev_pct    = (phev / total) * 100,
    hev_pct     = (hev  / total) * 100
  )

fso_long <- fso %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  select(year, ev_pct, phev_pct) %>%
  pivot_longer(-year, names_to = "type", values_to = "percentage") %>%
  mutate(
    type   = recode(type, "ev_pct" = "Pure Electric", "phev_pct" = "Hybrid"),
    source = "FSO (Official)"
  )

comparison_data_2 <- bind_rows(fso_long, sheds_md708_pop, sheds_history_pop) %>%
  mutate(
    source = factor(source, levels = c("FSO (Official)", "SHEDS (md_708)", "SHEDS (History)")),
    type   = factor(type,   levels = c("Hybrid", "Pure Electric", "Pure Electric (Main)", "Pure Electric (Secondary)"))
  )

comparison_totals_2 <- comparison_data %>%
  group_by(year, source) %>%
  summarise(total_pct = sum(percentage), .groups = "drop")

p_comparison <- comparison_data_2 %>%
  left_join(comparison_totals_2, by = c("year", "source")) %>%
  ggplot(aes(x = source, y = percentage, fill = type)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
            position = position_stack(vjust = 0.5),
            size = 3.5, color = "white", fontface = "bold") +
  geom_text(aes(y = total_pct, label = sprintf("%.1f%%", total_pct)),
            vjust = -0.5, size = 3.5, fontface = "bold", color = "black") +
  scale_fill_manual(values = c(
    "Hybrid"                    = "#FF9800",
    "Pure Electric"             = "#1B5E20",
    "Pure Electric (Main)"      = "#1B5E20",
    "Pure Electric (Secondary)" = "#81C784"
  )) +
  scale_y_continuous(
    limits = c(0, max(comparison_totals$total_pct) * 1.15),
    breaks = seq(0, 20, 2),
    expand = c(0, 0)
  ) +
  facet_wrap(~ year, nrow = 1) +
  labs(x = NULL, y = "Percentage of car owners (%)", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    axis.title.y       = element_text(size = 10, face = "bold"),
    axis.text.x        = element_text(size = 9, angle = 15, hjust = 1),
    legend.position    = "bottom",
    legend.text        = element_text(size = 16),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    strip.text         = element_text(face = "bold", size = 16)
  )

print(p_comparison)
save_plot(p_comparison, "bfs_vs_sheds_comparison_clean_csv", width = 14, height = 5)


```

Print comparison data:
```{r}
fso_counts <- fso %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  select(year, bev, phev, total) %>%
  pivot_longer(cols = c(bev, phev), names_to = "type", values_to = "count") %>%
  mutate(
    type       = recode(type, "bev" = "Pure Electric", "phev" = "Hybrid"),
    source     = "FSO (Official)",
    denominator = total
  ) %>%
  select(year, type, source, count, denominator)

# --- SHEDS md_708 counts ---
sheds_md708_counts <- md_708_summary %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  left_join(results_stata %>% select(year, n_car_owners), by = "year") %>%
  mutate(
    type        = "Pure Electric",
    source      = "SHEDS (md_708)",
    count       = n_ev_md708,
    denominator = n_car_owners
  ) %>%
  select(year, type, source, count, denominator)


sheds_history_counts <- results_stata %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(n_hybrid_total = n_hybrid_gas + n_hybrid_diesel) %>%
  select(year, n_ev_main, n_ev_secondary, n_hybrid_total, n_car_owners) %>%
  pivot_longer(cols = c(n_ev_main, n_ev_secondary, n_hybrid_total),
               names_to = "type", values_to = "count") %>%
  mutate(
    type        = recode(type,
                         "n_ev_main"      = "Pure Electric (Main)",
                         "n_ev_secondary" = "Pure Electric (Secondary)",
                         "n_hybrid_total" = "Hybrid"),
    source      = "SHEDS (History)",
    denominator = n_car_owners
  ) %>%
  select(year, type, source, count, denominator)


bind_rows(fso_counts, sheds_md708_counts, sheds_history_counts) %>%
  mutate(
    percentage = round((count / denominator) * 100, 2),
    source     = factor(source, levels = c("FSO (Official)", "SHEDS (md_708)", "SHEDS (History)")),
    type       = factor(type,   levels = c("Hybrid", "Pure Electric", "Pure Electric (Main)", "Pure Electric (Secondary)"))
  ) %>%
  arrange(year, source, type) %>%
  kable(
    col.names   = c("Year", "Type", "Source", "Count", "Total (denom.)", "Pct (%)"),
    align       = "rlrrrr",
    format.args = list(big.mark = ",")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  collapse_rows(columns = c(1, 3), valign = "top")
```


Make csv:
```{r}
sheds_counts <- results_stata %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(n_hybrid_total = n_hybrid_gas + n_hybrid_diesel) %>%
  left_join(
    md_708_summary %>% select(year, n_ev_md708),
    by = "year"
  ) %>%
  select(
    year,
    # denominators
    n_total,
    n_car_owners,
    n_car_owner_population,
    # md_708 source
    n_ev_md708,
    # history source
    n_ev_main,
    n_ev_secondary,
    n_ev_total,
    n_hybrid_gas,
    n_hybrid_diesel,
    n_hybrid_total
  )

print(sheds_counts)

write_csv(sheds_counts, "../sheds_vehicle_counts.csv")
```

check:
```{r}
stata_car_history %>%
  group_by(year_wave) %>%
  summarise(
    n_respondents    = n(),
    n_returning      = sum(id %in% stata_car_history$id[stata_car_history$year_wave < year_wave]),
    n_new            = n_respondents - n_returning
  )

stata_car_history %>%
  group_by(year_wave) %>%
  summarise(
    n_respondents = n(),
    n_returning   = sum(id %in% stata_car_history$id[stata_car_history$year_wave < year_wave]),
    n_new         = n_respondents - n_returning,
    n_car_owners  = sum(mob2_1 >= 1 & mob2_1 < 90, na.rm = TRUE),
    n_no_car      = n_respondents - n_car_owners,
    pct_car_owners = round(n_car_owners / n_respondents * 100, 1)
  ) %>%
  left_join(
    results_stata %>% select(year, n_car_owners) %>% rename(n_car_owners_stata = n_car_owners),
    by = c("year_wave" = "year")
  ) %>%
  mutate(match = n_car_owners == n_car_owners_stata) %>%
  kable(
    col.names = c("Year", "N Respondents", "N Returning", "N New",
                  "N Car Owners", "N No Car", "% Car Owners",
                  "N Car Owners (results_stata)", "Match"),
    align = "r",
    format.args = list(big.mark = ",")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

