---
title: "Longitudinal exploration of cars"
output: html_notebook
---

```{r}
library(haven)
library(tidyverse)
library(skimr)
library(corrplot)
library(scales)
library(ggthemes)
library(kableExtra)

current_dir <- getwd()
source(file.path(current_dir,"src","R","utils.R"))

years <- c(2016, 2017, 2018, 2019, 2020, 2021, 2023, 2025)

waves <- list()

data_path <- "/home/amy/tmp/sheds_data/files/SHEDS"

waves <- list()
for (year in years) {
  file_path <- paste0("/home/amy/tmp/sheds_data/files/SHEDS", year, ".sav")
  wave_name <- as.character(year)
  waves[[wave_name]] <- read_clean_sheds(file_path)
}

all_ids <- waves %>%
  map_df(~ select(.x, id)) %>%
  pull(id) %>%
  unique()

cat("\nTotal unique IDs across all waves:", length(all_ids), "\n")

cat("\n=== Summary ===\n")
cat("Total respondents (all waves):", sum(map_int(waves, nrow)), "\n")
cat("Total unique IDs:", length(all_ids), "\n")
cat("Duplicate appearances:", sum(map_int(waves, nrow)) - length(all_ids), "\n")

id_counts <- waves %>%
  map_df(~ select(.x, id), .id = "wave") %>%
  count(id) %>%
  arrange(desc(n))

cat("\nDistribution of appearances:\n")
print(table(id_counts$n))
```

EV ownership over time using md_708 and history building:
```{r}
library(tidyverse)
library(scales)
library(knitr)

md_708_summary <- bind_rows(
  lapply(names(waves), function(wave_name) {
    df <- waves[[wave_name]]
    tibble(
      year           = as.integer(wave_name),
      n_total        = nrow(df),
      n_population   = if ("md_hhgr" %in% names(df)) sum(df$md_hhgr[df$mob2_1 >= 1 & df$mob2_1 < 90], na.rm = TRUE) else NA_integer_,
      n_ev_md708     = if ("md_708" %in% names(df)) sum(df$md_708 == 1, na.rm = TRUE) else NA_integer_,
      n_ev_missing   = if ("md_708" %in% names(df)) sum(is.na(df$md_708)) else NA_integer_,
      ev_rate_md708  = n_ev_md708 / n_total
    )
  })
) %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025))

md_708_summary %>%
  mutate(ev_rate_md708 = percent(ev_rate_md708, accuracy = 0.01)) %>%
  kable(
    col.names  = c("Year", "N Total", "N Pop.", "N EV (md_708)", "N Missing", "EV Rate"),
    align       = "r",
    format.args = list(big.mark = ",")
  ) %>%
  print()

stata_car_history <- build_car_history(waves)


results_stata <- bind_rows(
  analyze_ev_ownership_data(stata_car_history, 2019, waves),
  analyze_ev_ownership_data(stata_car_history, 2020, waves ),
  analyze_ev_ownership_data(stata_car_history, 2021, waves),
  analyze_ev_ownership_data(stata_car_history, 2023, waves),
  analyze_ev_ownership_data(stata_car_history, 2025, waves)
)

results_stata %>%
  select(year, n_total, n_car_owner_population, n_car_owners, n_ev_main, n_ev_secondary,
         n_ev_total, n_hybrid_gas, n_hybrid_diesel, ev_rate_car_owners) %>%
  mutate(ev_rate_car_owners = percent(ev_rate_car_owners, accuracy = 0.01)) %>%
  kable(
    col.names = c("Year", "N", "Pop.", "Car Own.", "EV Main", "EV 2nd",
                  "EV Total", "Hybrid Gas", "Hybrid Dies.", "EV Rate"),
    align = "r",
    format.args = list(big.mark = ",")
  ) %>%
  print()
```


Here we want to see who are the people which have md_708 and who have ev in the car history based on the ids what are the differences.
Did people responded differently to each questions? or maybe they didnt say the changed vehicles can we track ids from both dataframes and check
why it differs?

Diagnostic: Who is in md_708 vs car history - and why do they differ?
```{r}
# For each year, build a per-respondent comparison between the two EV sources
diagnose_ev_discrepancy <- function(yr) {
  raw <- waves[[as.character(yr)]] %>%
    mutate(across(everything(), as.vector)) %>%
    select(id,
           any_of("md_708"),
           any_of("mob2_1"),
           any_of("mob3_3"),
           any_of("mob3_change"),
           any_of("mob2_e")) %>%
    mutate(
      md_708      = ifelse("md_708" %in% names(.) & !is.na(md_708) & md_708 %in% c(-1,-2), NA_real_, as.numeric(md_708)),
      mob3_3      = ifelse(mob3_3 %in% c(-2,-1), NA_real_, as.numeric(mob3_3)),
      mob3_change = ifelse("mob3_change" %in% names(.) & !is.na(mob3_change) & mob3_change %in% c(-1,-2), NA_real_, as.numeric(mob3_change)),
      mob2_e      = ifelse("mob2_e" %in% names(.) & !is.na(mob2_e) & mob2_e %in% c(-1,-2), NA_real_, as.numeric(mob2_e))
    )

hist_yr <- stata_car_history %>%
    filter(year_wave == yr) %>%
    mutate(has_ev = as.integer(mob3_3_filled == 8 | mob2_e_filled == 1)) %>%
    select(id, mob3_3_filled, mob2_e_filled, has_ev)

  joined <- raw %>%
    left_join(hist_yr, by = "id") %>%
    mutate(
      ev_md708   = as.integer(!is.na(md_708) & md_708 == 1),
      ev_history = as.integer(!is.na(has_ev) & has_ev == 1),
      group = case_when(
        ev_md708 == 1 & ev_history == 1  ~ "Both agree: EV",
        ev_md708 == 0 & ev_history == 0  ~ "Both agree: No EV",
        ev_md708 == 1 & ev_history == 0  ~ "md_708 only (history missed)",
        ev_md708 == 0 & ev_history == 1  ~ "History only (md_708 missed)",
        is.na(ev_md708) & ev_history == 1 ~ "History only (md_708 NA)",
        ev_md708 == 1 & is.na(ev_history) ~ "md_708 only (not in history)",
        TRUE                                ~ "Both NA / unclear"
      )
    )
  joined
}

diag_table <- bind_rows(
  lapply(c(2019, 2020, 2021, 2023, 2025), function(yr) {
    d <- diagnose_ev_discrepancy(yr)
    n_total <- nrow(waves[[as.character(yr)]])
    bind_rows(
      d %>% count(group) %>% mutate(year = yr, n_total = n_total),
      tibble(
        group = c(
          "TOTAL: EV in md_708 (all IDs)",
          "TOTAL: EV in history (all IDs)",
          "TOTAL: EV in either source (union)",
          "TOTAL: EV in both sources (intersection)",
          "TOTAL: md_708 missing & EV via secondary only",
          "TOTAL: md_708 missing & EV via main only"
        ),
        n = c(
          sum(d$ev_md708   == 1, na.rm = TRUE),
          sum(d$ev_history == 1, na.rm = TRUE),
          sum(d$ev_md708 == 1 | d$ev_history == 1, na.rm = TRUE),
          sum(d$ev_md708 == 1 & d$ev_history == 1, na.rm = TRUE),
          sum(is.na(d$md_708) & d$mob2_e_filled == 1 & d$mob3_3_filled != 8, na.rm = TRUE),
          sum(is.na(d$md_708) & d$mob3_3_filled == 8, na.rm = TRUE)
        ),
        year    = yr,
        n_total = n_total
      )
    )
  })
) %>%
  mutate(label = sprintf("%d (%.1f%%)", n, n / n_total * 100)) %>%
  select(group, year, label) %>%
  pivot_wider(names_from = year, values_from = label, values_fill = "0 (0.0%)") %>%
  arrange(group)

total_rows <- which(str_detect(diag_table$group, "TOTAL"))

diag_table %>%
  kable(caption = "Agreement between md_708 and car history by year (joined on IDs)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  row_spec(total_rows, bold = TRUE, background = "#eeeeee")
```


People history says EV but not md708:
```{r}
# People md_708 says has EV but history missed
diag_2025 <- diagnose_ev_discrepancy(2025)
md708_only <- diag_2025 %>%
  filter(group == "md_708 only (history missed)") %>%
  select(id, md_708, mob3_3, mob3_3_filled, mob3_change, mob2_e, mob2_e_filled)

cat("\n--- md_708 says EV, history missed:", nrow(md708_only), "respondents ---\n")
cat("Did they report a car change (mob3_change)?\n")
print(table(md708_only$mob3_change, useNA = "always"))
cat("\nRaw mob3_3 engine type in this wave:\n")
print(table(md708_only$mob3_3, useNA = "always"))
cat("\nFilled mob3_3 from history:\n")
print(table(md708_only$mob3_3_filled, useNA = "always"))
cat("\nNA mob3_3 (never reported engine type):", sum(is.na(md708_only$mob3_3)), "\n")
cat("Non-EV engine type in mob3_3 (possible misclassification):",
    sum(!is.na(md708_only$mob3_3) & !md708_only$mob3_3 %in% 8, na.rm = TRUE), "\n")
```

```{r}
# People history says EV but md_708 missed
history_only <- diag_2025 %>%
  filter(group == "History only (md_708 missed)") %>%
  select(id, md_708, mob3_3, mob3_3_filled, mob3_change, mob2_e, mob2_e_filled)

cat("\n--- History says EV, md_708 missed:", nrow(history_only), "respondents ---\n")
cat("Raw md_708 for these respondents:\n")
print(table(history_only$md_708, useNA = "always"))
cat("\nFilled engine type (mob3_3_filled):\n")
print(table(history_only$mob3_3_filled, useNA = "always"))
cat("\nEV via secondary vehicle (mob2_e_filled):\n")
print(table(history_only$mob2_e_filled, useNA = "always"))
```


Discrepancy over time all years:
```{r}
# Track discrepancy across all years
discrepancy_over_time <- bind_rows(
  lapply(c(2019, 2020, 2021, 2023, 2025), function(yr) {
    d <- diagnose_ev_discrepancy(yr)
    d %>% count(group) %>% mutate(year = yr, pct = n / sum(n) * 100)
  })
)

discrepancy_over_time %>%
  filter(!group %in% c("Both agree: No EV", "Both NA / unclear")) %>%
  ggplot(aes(x = year, y = pct, colour = group, group = group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(aes(label = sprintf("%.1f%%", pct)), vjust = -0.8, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c(
    "Both agree: EV"                 = "#1B5E20",
    "md_708 only (history missed)"   = "#FF9800",
    "History only (md_708 missed)"   = "#1565C0",
    "History only (md_708 NA)"       = "#90CAF9",
    "md_708 only (not in history)"   = "#FFB74D"
  )) +
  scale_y_continuous(labels = label_percent(scale = 1), expand = expansion(mult = c(0, 0.2))) +
  labs(
    title    = "Agreement Between md_708 and Car History EV Classification",
    subtitle = "% of all respondents per wave",
    x = NULL, y = "% of respondents", colour = NULL
  ) +
  theme_publication(base_size = 12)
```

Plots for household:

```{r}
bfs_percentage <- read_csv("/home/amy/tmp/repos/sheds_data_analysis/gr-e-11.03.02.01-03.csv", show_col_types = FALSE) %>%
  rename(year = `Unnamed: 0`) %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(
    hybrid_total = `conventional hybrid` + `plug-in hybrid`,
    total_vehicles = petrol + diesel + `conventional hybrid` + `plug-in hybrid` + electric + other,
    ev_pct = (electric / total_vehicles) * 100,
    hybrid_pct = (hybrid_total / total_vehicles) * 100
  ) %>%
  select(year, ev_pct, hybrid_pct) %>%
  pivot_longer(-year, names_to = "type", values_to = "percentage") %>%
  mutate(
    type = recode(type, "ev_pct" = "Pure Electric", "hybrid_pct" = "Hybrid"),
    source = "FSO (Official)"
  )

sheds_md708 <- md_708_summary %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(
    percentage = (n_ev_md708 / n_total) * 100,
    type = "Pure Electric",
    source = "SHEDS (md_708)"
  ) %>%
  select(year, type, percentage, source)

sheds_history <- results_stata %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(n_hybrid_total = n_hybrid_gas + n_hybrid_diesel) %>%
  select(year, n_ev_main, n_ev_secondary, n_hybrid_total, n_total) %>%
  pivot_longer(cols = c(n_ev_main, n_ev_secondary, n_hybrid_total),
               names_to = "type", values_to = "count") %>%
  mutate(
    percentage = (count / n_total) * 100,
    type = recode(type,
                  "n_ev_main" = "Pure Electric (Main)",
                  "n_ev_secondary" = "Pure Electric (Secondary)",
                  "n_hybrid_total" = "Hybrid"),
    source = "SHEDS (History)"
  ) %>%
  select(year, type, percentage, source)

comparison_data_household <- bind_rows(bfs_percentage, sheds_md708, sheds_history) %>%
  mutate(
    source = factor(source, levels = c("FSO (Official)", "SHEDS (md_708)", "SHEDS (History)")),
    type = factor(type, levels = c("Hybrid", "Pure Electric", "Pure Electric (Main)", "Pure Electric (Secondary)"))
  )

comparison_totals_household <- comparison_data_household %>%
  group_by(year, source) %>%
  summarise(total_pct = sum(percentage), .groups = "drop")

p_comparison_household <- comparison_data_household %>%
  left_join(comparison_totals_household, by = c("year", "source")) %>%
  ggplot(aes(x = source, y = percentage, fill = type)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
            position = position_stack(vjust = 0.5),
            size = 3.5, color = "white", fontface = "bold") +
  geom_text(aes(y = total_pct, label = sprintf("%.1f%%", total_pct)),
            vjust = -0.5, size = 3.5, fontface = "bold", color = "black") +
  scale_fill_manual(
    values = c(
      "Hybrid" = "#FF9800",
      "Pure Electric" = "#1B5E20",
      "Pure Electric (Main)" = "#1B5E20",
      "Pure Electric (Secondary)" = "#81C784"
    )
  ) +
  scale_y_continuous(limits = c(0, max(comparison_totals_household$total_pct) * 1.15),
                     breaks = seq(0, 20, 2),
                     expand = c(0, 0)) +
  facet_wrap(~ year, nrow = 1) +
  labs(
    x = NULL,
    y = "Percentage (%)",
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 16),
    axis.title.y = element_text(size = 10, face = "bold"),
    axis.text.x = element_text(size = 9, angle = 15, hjust = 1),
    legend.position = "bottoub/api/dam/assets/23366958/master",
    legend.text = element_text(size = 16),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 16)
  )

print(p_comparison_household)
save_plot(p_comparison_household, "bfs_vs_sheds_comparison_household", width = 14, height = 5)
```


Plots with the population so divided by the members of household:
```{r}
bfs_percentage <- read_csv("/home/amy/tmp/repos/sheds_data_analysis/gr-e-11.03.02.01-03.csv", show_col_types = FALSE) %>%
  rename(year = `Unnamed: 0`) %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(
    hybrid_total = `conventional hybrid` + `plug-in hybrid`,
    total_vehicles = petrol + diesel + `conventional hybrid` + `plug-in hybrid` + electric + other,
    ev_pct = (electric / total_vehicles) * 100,
    hybrid_pct = (hybrid_total / total_vehicles) * 100
  ) %>%
  select(year, ev_pct, hybrid_pct) %>%
  pivot_longer(-year, names_to = "type", values_to = "percentage") %>%
  mutate(
    type = recode(type, "ev_pct" = "Pure Electric", "hybrid_pct" = "Hybrid"),
    source = "FSO (Official)"
  )

sheds_md708_pop <- md_708_summary %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  #left_join(results_stata %>% select(year, n_population), by = "year") %>%
  mutate(
    percentage = (n_ev_md708 / n_population) * 100,
    type = "Pure Electric",
    source = "SHEDS (md_708)"
  ) %>%
  select(year, type, percentage, source)

sheds_history_pop <- results_stata %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(n_hybrid_total = n_hybrid_gas + n_hybrid_diesel) %>%
  select(year, n_ev_main, n_ev_secondary, n_hybrid_total, n_car_owner_population) %>%
  pivot_longer(cols = c(n_ev_main, n_ev_secondary, n_hybrid_total),
               names_to = "type", values_to = "count") %>%
  mutate(
    percentage = (count / n_car_owner_population) * 100,
    type = recode(type,
                  "n_ev_main" = "Pure Electric (Main)",
                  "n_ev_secondary" = "Pure Electric (Secondary)",
                  "n_hybrid_total" = "Hybrid"),
    source = "SHEDS (History)"
  ) %>%
  select(year, type, percentage, source)

comparison_data_population <- bind_rows(bfs_percentage, sheds_md708_pop, sheds_history_pop) %>%
  mutate(
    source = factor(source, levels = c("FSO (Official)", "SHEDS (md_708)", "SHEDS (History)")),
    type = factor(type, levels = c("Hybrid", "Pure Electric", "Pure Electric (Main)", "Pure Electric (Secondary)"))
  )

comparison_totals_population <- comparison_data_population %>%
  group_by(year, source) %>%
  summarise(total_pct = sum(percentage), .groups = "drop")

p_comparison_population <- comparison_data_population %>%
  left_join(comparison_totals_population, by = c("year", "source")) %>%
  ggplot(aes(x = source, y = percentage, fill = type)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
            position = position_stack(vjust = 0.5),
            size = 3.5, color = "white", fontface = "bold") +
  geom_text(aes(y = total_pct, label = sprintf("%.1f%%", total_pct)),
            vjust = -0.5, size = 3.5, fontface = "bold", color = "black") +
  scale_fill_manual(
    values = c(
      "Hybrid" = "#FF9800",
      "Pure Electric" = "#1B5E20",
      "Pure Electric (Main)" = "#1B5E20",
      "Pure Electric (Secondary)" = "#81C784"
    )
  ) +
  scale_y_continuous(limits = c(0, max(comparison_totals_population$total_pct) * 1.15),
                     breaks = seq(0, 20, 2),
                     expand = c(0, 0)) +
  facet_wrap(~ year, nrow = 1) +
  labs(
    x = NULL,
    y = "Percentage (%)",
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 16),
    axis.title.y = element_text(size = 10, face = "bold"),
    axis.text.x = element_text(size = 9, angle = 15, hjust = 1),
    legend.position = "bottom",
    legend.text = element_text(size = 16),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 16)
  )

print(p_comparison_population)
save_plot(p_comparison_population, "bfs_vs_sheds_comparison_population", width = 14, height = 5)
```



Who are the people which have EV md_708 and people who are in the SHEDS history whats the difference?
```{r}
waves[["2025"]] %>%
  select(mob3_3) %>%
  mutate(mob3_3_raw = haven::as_factor(mob3_3)) %>%
  count(mob3_3, mob3_3_raw) %>%
  arrange(mob3_3)
```

Population just ev:
```{r}
# --- BFS: EV only ---
bfs_percentage_ev <- read_csv("/home/amy/tmp/repos/sheds_data_analysis/gr-e-11.03.02.01-03.csv",
                               show_col_types = FALSE) %>%
  rename(year = `Unnamed: 0`) %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(
    total_vehicles = petrol + diesel + `conventional hybrid` + `plug-in hybrid` + electric + other,
    ev_pct         = (electric / total_vehicles) * 100
  ) %>%
  select(year, ev_pct) %>%
  pivot_longer(-year, names_to = "type", values_to = "percentage") %>%
  mutate(
    type   = recode(type, "ev_pct" = "Pure Electric"),
    source = "FSO (Official)"
  )

# --- SHEDS md_708: EV only (unchanged, already no hybrid) ---
sheds_md708_pop_ev <- md_708_summary %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(
    percentage = (n_ev_md708 / n_population) * 100,
    type       = "Pure Electric",
    source     = "SHEDS (md_708)"
  ) %>%
  select(year, type, percentage, source)

# --- SHEDS history: EV only (drop n_hybrid_total) ---
sheds_history_pop_ev <- results_stata %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  select(year, n_ev_main, n_ev_secondary, n_car_owner_population) %>%
  pivot_longer(cols = c(n_ev_main, n_ev_secondary),
               names_to = "type", values_to = "count") %>%
  mutate(
    percentage = (count / n_car_owner_population) * 100,
    type       = recode(type,
                        "n_ev_main"       = "Pure Electric (Main)",
                        "n_ev_secondary"  = "Pure Electric (Secondary)"),
    source     = "SHEDS (History)"
  ) %>%
  select(year, type, percentage, source)

# --- Combine & plot ---
comparison_data_population_ev <- bind_rows(
  bfs_percentage_ev, sheds_md708_pop_ev, sheds_history_pop_ev
) %>%
  mutate(
    source = factor(source, levels = c("FSO (Official)", "SHEDS (md_708)", "SHEDS (History)")),
    type   = factor(type,   levels = c("Pure Electric", "Pure Electric (Main)", "Pure Electric (Secondary)"))
  )

comparison_totals_population_ev <- comparison_data_population_ev %>%
  group_by(year, source) %>%
  summarise(total_pct = sum(percentage), .groups = "drop")

p_comparison_population_ev <- comparison_data_population_ev %>%
  left_join(comparison_totals_population_ev, by = c("year", "source")) %>%
  ggplot(aes(x = source, y = percentage, fill = type)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
            position = position_stack(vjust = 0.5),
            size = 3.5, color = "white", fontface = "bold") +
  geom_text(aes(y = total_pct, label = sprintf("%.1f%%", total_pct)),
            vjust = -0.5, size = 3.5, fontface = "bold", color = "black") +
  scale_fill_manual(
    values = c(
      "Pure Electric"             = "#1B5E20",
      "Pure Electric (Main)"      = "#1B5E20",
      "Pure Electric (Secondary)" = "#81C784"
    )
  ) +
  scale_y_continuous(
    limits = c(0, max(comparison_totals_population_ev$total_pct) * 1.15),
    breaks = seq(0, 20, 2),
    expand = c(0, 0)
  ) +
  facet_wrap(~ year, nrow = 1) +
  labs(x = NULL, y = "Percentage (%)", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title        = element_text(face = "bold", size = 16),
    axis.title.y      = element_text(size = 10, face = "bold"),
    axis.text.x       = element_text(size = 9, angle = 15, hjust = 1),
    legend.position   = "bottom",
    legend.text       = element_text(size = 16),
    panel.grid.major.x = element_blank(),
    panel.grid.minor  = element_blank(),
    strip.text        = element_text(face = "bold", size = 16)
  )

print(p_comparison_population_ev)
save_plot(p_comparison_population_ev, "bfs_vs_sheds_comparison_population_ev", width = 14, height = 5)
```


Household ev new:

```{r}
# --- BFS: EV only ---
bfs_percentage_ev <- read_csv("/home/amy/tmp/repos/sheds_data_analysis/gr-e-11.03.02.01-03.csv",
                               show_col_types = FALSE) %>%
  rename(year = `Unnamed: 0`) %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(
    total_vehicles = petrol + diesel + `conventional hybrid` + `plug-in hybrid` + electric + other,
    ev_pct         = (electric / total_vehicles) * 100
  ) %>%
  select(year, ev_pct) %>%
  pivot_longer(-year, names_to = "type", values_to = "percentage") %>%
  mutate(
    type   = recode(type, "ev_pct" = "Pure Electric"),
    source = "FSO (Official)"
  )

# --- SHEDS md_708: EV only (unchanged, already no hybrid) ---
sheds_md708_ev <- md_708_summary %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(
    percentage = (n_ev_md708 / n_total) * 100,
    type       = "Pure Electric",
    source     = "SHEDS (md_708)"
  ) %>%
  select(year, type, percentage, source)

# --- SHEDS history: EV only (drop n_hybrid_total) ---
sheds_history_ev <- results_stata %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  select(year, n_ev_main, n_ev_secondary, n_total) %>%
  pivot_longer(cols = c(n_ev_main, n_ev_secondary),
               names_to = "type", values_to = "count") %>%
  mutate(
    percentage = (count / n_total) * 100,
    type       = recode(type,
                        "n_ev_main"      = "Pure Electric (Main)",
                        "n_ev_secondary" = "Pure Electric (Secondary)"),
    source     = "SHEDS (History)"
  ) %>%
  select(year, type, percentage, source)

# --- Combine & plot ---
comparison_data_household_ev <- bind_rows(
  bfs_percentage_ev, sheds_md708_ev, sheds_history_ev
) %>%
  mutate(
    source = factor(source, levels = c("FSO (Official)", "SHEDS (md_708)", "SHEDS (History)")),
    type   = factor(type,   levels = c("Pure Electric", "Pure Electric (Main)", "Pure Electric (Secondary)"))
  )

comparison_totals_household_ev <- comparison_data_household_ev %>%
  group_by(year, source) %>%
  summarise(total_pct = sum(percentage), .groups = "drop")

p_comparison_household_ev <- comparison_data_household_ev %>%
  left_join(comparison_totals_household_ev, by = c("year", "source")) %>%
  ggplot(aes(x = source, y = percentage, fill = type)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
            position = position_stack(vjust = 0.5),
            size = 3.5, color = "white", fontface = "bold") +
  geom_text(aes(y = total_pct, label = sprintf("%.1f%%", total_pct)),
            vjust = -0.5, size = 3.5, fontface = "bold", color = "black") +
  scale_fill_manual(
    values = c(
      "Pure Electric"             = "#1B5E20",
      "Pure Electric (Main)"      = "#1B5E20",
      "Pure Electric (Secondary)" = "#81C784"
    )
  ) +
  scale_y_continuous(
    limits = c(0, max(comparison_totals_household_ev$total_pct) * 1.15),
    breaks = seq(0, 20, 2),
    expand = c(0, 0)
  ) +
  facet_wrap(~ year, nrow = 1) +
  labs(x = NULL, y = "Percentage (%)", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title         = element_text(face = "bold", size = 16),
    axis.title.y       = element_text(size = 10, face = "bold"),
    axis.text.x        = element_text(size = 9, angle = 15, hjust = 1),
    legend.position    = "bottom",
    legend.text        = element_text(size = 16),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    strip.text         = element_text(face = "bold", size = 16)
  )

print(p_comparison_household_ev)
save_plot(p_comparison_household_ev, "bfs_vs_sheds_comparison_household_ev", width = 14, height = 5)
```

New data graph:
```{r}
bfs_percentage <- read_csv("/home/amy/tmp/repos/sheds_data_analysis/gr-e-11.03.02.01-03_new.csv",
                           show_col_types = FALSE) %>%
  rename(
    year    = `Unnamed: 0`,
    petrol  = `petrol (incl. hybrid electric vehicle, HEV)`,
    diesel  = `diesel (incl. hybrid electric vehicle, HEV)`,
    phev    = `plug-in hybrid electric vehicle (PHEV)`,
    bev     = `battery electric vehicle (BEV)`
  ) %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(
    hybrid_total   = phev,
    total_vehicles = petrol + diesel + phev + bev + other,
    ev_pct         = (bev  / total_vehicles) * 100,
    hybrid_pct     = (phev / total_vehicles) * 100
  )

print(bfs_percentage)
```

EV new population:
```{r}
# --- BFS new file: EV only ---
bfs_percentage_new_ev <- read_csv("/home/amy/tmp/repos/sheds_data_analysis/gr-e-11.03.02.01-03_new.csv",
                                   show_col_types = FALSE) %>%
  rename(
    year   = `Unnamed: 0`,
    petrol = `petrol (incl. hybrid electric vehicle, HEV)`,
    diesel = `diesel (incl. hybrid electric vehicle, HEV)`,
    phev   = `plug-in hybrid electric vehicle (PHEV)`,
    bev    = `battery electric vehicle (BEV)`
  ) %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(
    total_vehicles = petrol + diesel + phev + bev + other,
    ev_pct         = (bev / total_vehicles) * 100
  ) %>%
  select(year, ev_pct) %>%
  pivot_longer(-year, names_to = "type", values_to = "percentage") %>%
  mutate(
    type   = recode(type, "ev_pct" = "Pure Electric"),
    source = "FSO (Official)"
  )

# --- SHEDS md_708: EV only ---
sheds_md708_pop_ev <- md_708_summary %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  mutate(
    percentage = (n_ev_md708 / n_population) * 100,
    type       = "Pure Electric",
    source     = "SHEDS (md_708)"
  ) %>%
  select(year, type, percentage, source)

# --- SHEDS history: EV only ---
sheds_history_pop_ev <- results_stata %>%
  filter(year %in% c(2019, 2020, 2021, 2023, 2025)) %>%
  select(year, n_ev_main, n_ev_secondary, n_car_owner_population) %>%
  pivot_longer(cols = c(n_ev_main, n_ev_secondary),
               names_to = "type", values_to = "count") %>%
  mutate(
    percentage = (count / n_car_owner_population) * 100,
    type       = recode(type,
                        "n_ev_main"      = "Pure Electric (Main)",
                        "n_ev_secondary" = "Pure Electric (Secondary)"),
    source     = "SHEDS (History)"
  ) %>%
  select(year, type, percentage, source)

# --- Combine & plot ---
comparison_data_population_new_ev <- bind_rows(
  bfs_percentage_new_ev, sheds_md708_pop_ev, sheds_history_pop_ev
) %>%
  mutate(
    source = factor(source, levels = c("FSO (Official)", "SHEDS (md_708)", "SHEDS (History)")),
    type   = factor(type,   levels = c("Pure Electric", "Pure Electric (Main)", "Pure Electric (Secondary)"))
  )

comparison_totals_population_new_ev <- comparison_data_population_new_ev %>%
  group_by(year, source) %>%
  summarise(total_pct = sum(percentage), .groups = "drop")

p_comparison_population_new_ev <- comparison_data_population_new_ev %>%
  left_join(comparison_totals_population_new_ev, by = c("year", "source")) %>%
  ggplot(aes(x = source, y = percentage, fill = type)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
            position = position_stack(vjust = 0.5),
            size = 3.5, color = "white", fontface = "bold") +
  geom_text(aes(y = total_pct, label = sprintf("%.1f%%", total_pct)),
            vjust = -0.5, size = 3.5, fontface = "bold", color = "black") +
  scale_fill_manual(
    values = c(
      "Pure Electric"             = "#1B5E20",
      "Pure Electric (Main)"      = "#1B5E20",
      "Pure Electric (Secondary)" = "#81C784"
    )
  ) +
  scale_y_continuous(
    limits = c(0, max(comparison_totals_population_new_ev$total_pct) * 1.15),
    breaks = seq(0, 20, 2),
    expand = c(0, 0)
  ) +
  facet_wrap(~ year, nrow = 1) +
  labs(x = NULL, y = "Percentage (%)", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title         = element_text(face = "bold", size = 16),
    axis.title.y       = element_text(size = 10, face = "bold"),
    axis.text.x        = element_text(size = 9, angle = 15, hjust = 1),
    legend.position    = "bottom",
    legend.text        = element_text(size = 16),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    strip.text         = element_text(face = "bold", size = 16)
  )

print(p_comparison_population_new_ev)
save_plot(p_comparison_population_new_ev, "bfs_vs_sheds_comparison_population_new_ev", width = 14, height = 5)
```
